Parsing Postfix Log Files

### 2009-04-22 doc/logparser-acronyms.tex
Corrections based on proofreading to the preface and chapters 1-3.
Corrections after proofreading chapters 6 & 7 and the appendices.

\newglossaryentry{DNSBL}{
    type=\acronymtype,
    name={DNSBL},
    description={DNS Blacklist},
    first={DNS Blacklist (DNSBL)}
}
\newglossaryentry{DNSBL glossary}{
    name={DNS Blacklist},
    description={
        A DNS Blacklist is a simple collaborative anti-spam technique used
        to reject or penalise mail sent from mail servers believed to be
        the source of large volumes of spam.  The criteria used when
        deciding if an IP address should be included on a DNSBL vary widely
        between DNSBLs, so before using one it is very important to check
        their listing policies.  To use a DNSBL the mail server makes a DNS
        request incorporating the IP address of the client; if the
        requested hostname exists the client is on the DNSBL, and the mail
        server can decide what course of action to take.
    }
}

\newglossaryentry{API}{
    type=\acronymtype,
    name={API},
    description={Application Programming Interface},
    first={Application Programming Interface (API)}
}
\newglossaryentry{API glossary}{
    name={Application Programming Interface},
    description={
        One of the fundamental concepts when writing programs is the reuse
        of existing code, so that each new program does not reinvent
        existing wheels.  An API defines the interface provided to the user
        of the existing code, and acts as a contract between the user and
        the provider: if the user adheres to the API the provider
        guarantees it will work, but is free to change the implementation
        if the API is preserved.
    }
}

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newcommand{\SMTPglossaryDescription}[0]{%
    \acronym{SMTP} is the protocol used for transferring mail between the
    sending and receiving \acronym{MTA}\@.  It is a simple, human readable,
    plain text protocol, making it quite simple to test and debug problems
    with it.  A detailed description of SMTP is beyond the scope of this
    thesis: the original protocol definition is in
    \acronym{RFC}~821~\cite{RFC821}, which was updated in
    \acronym{RFC}~2821~\cite{RFC2821}.%
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        \SMTPglossaryDescription{}
    }
}

\newglossaryentry{LMTP}{
    type=\acronymtype,
    name={LMTP},
    description={Local Mail Transfer Protocol},
    first={Local Mail Transfer Protocol (LMTP)}
}
\newglossaryentry{LMTP glossary}{
    name={Local Mail Transfer Protocol},
    description={
        LMTP is a protocol derived from SMTP that removes the need for the
        server to maintain a mail delivery queue, instead relying on the
        client to maintain it.  Typically the client is an MTA, and the
        server a delivery agent or a mail store.  Full details are
        available in~\cite{lmtp-rfc}.
    }
}

\newglossaryentry{MTA}{
    type=\acronymtype,
    name={MTA},
    description={Mail Transfer Agent},
    first={Mail Transfer Agent (MTA)}
}
\newglossaryentry{MTA glossary}{
    name={Mail Transfer Agent},
    description={
        A Mail Transfer Agent sends and receives mail via SMTP\@.  Users
        submit mail to an MTA via their mail client (e.g.\ Microsoft
        Outlook, Thunderbird, webmail services); the sending MTA transfers
        the mail to the receiving MTA (which may in turn forward it to
        another MTA).
    }
}

\newglossaryentry{RFC}{
    type=\acronymtype,
    name={RFC},
    description={Request For Comments},
    first={Request For Comments (RFC)}
}
\newglossaryentry{RFC glossary}{
    name={Request For Comments},
    description={
        The Request For Comments series is a series of proposals defining
        various protocols and file formats, e.g.\ SMTP\@.  The name is
        somewhat misleading nowadays: initially the authors were asking for
        peer review, but these documents are now the de facto standards the
        Internet runs on.
    }
}

@@

\newglossaryentry{FQDN}{
    type=\acronymtype,
    name={FQDN},
    description={Fully Qualified Domain Name},
    first={Fully Qualified Domain Name (FQDN)}
}
\newglossaryentry{FQDN glossary}{
    name={Fully Qualified Domain Name},
    description={
        An FQDN is a hostname plus domain name, e.g.\ \newline{}
            example.com     is a domain name          \newline{}
            www             is a hostname             \newline{}
            www.example.com is a FQDN
    }
}

\newglossaryentry{pid}{
    type=\acronymtype,
    name={pid},
    description={Process Identifier},
    first={Process Identifier (pid)}
}
\newglossaryentry{pid glossary}{
    name={Process Identifier},
    description={
        There may be multiple copies of any program executing at any one
        time, so the program's name is not suitable as a distinguishing
        identifier; instead each process executing is given a pid that is
        guaranteed to be unique for the lifetime of the process.  Once the
        process has completed, the pid may be reused, as they are drawn
        from a finite pool.
    }
}

\newglossaryentry{UCE}{
    type=\acronymtype,
    name={UCE},
    description={Unsolicited Commercial Email},
    first={Unsolicited Commercial Email (UCE)}
}
\newglossaryentry{UCE glossary}{
    name={Unsolicited Commercial Email},
    description={
        UCE is a more restrictive definition of spam than most people use:
        it only covers mail that is explicitly commercial, thus excluding
        viruses, Bayesian poisoning mails, backscatter\glsadd{Backscatter},
        and those annoying chain letters you get from friends.
    }
}

\newglossaryentry{ESMTP}{
    type=\acronymtype,
    name={ESMTP},
    description={Extended SMTP},
    first={Extended SMTP (ESMTP)}
}
\newglossaryentry{ESMTP glossary}{
    name={Extended SMTP},
    description={
        \acronym{ESMTP}~\cite{RFC1869}, provides a flexible mechanism for
        \acronym{SMTP} to be extended with new functionality, allowing new
        features to be tested without having to be included in the standard
        protocol.  \acronym{ESMTP} is backwards compatible with
        \acronym{SMTP}: \acronym{ESMTP} clients and servers can interact
        with \acronym{SMTP} clients and servers without difficulty.
    }
}

\newglossaryentry{IP}{
    type=\acronymtype,
    name={IP},
    description={Internet Protocol},
    first={Internet Protocol (IP)}
}

@@

\newglossaryentry{FA}{
    type=\acronymtype,
    name={FA},
    description={Finite Automata},
    first={Finite Automata (FA)}
}
\newglossaryentry{FA glossary}{
    name={Finite Automata},
    description={
        Finite automata are computing devices that accept or recognize
        regular languages.  They lack any form of storage, so although they
        can recognise languages such as ${(ab)}^{*}$ or
        $a^{*}{(b^{*}|c)}b$, they cannot count and so cannot recognise
        languages such as $a^{n}b^{n}$.
    }
}

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        A CFG is a grammar in which every transition or production takes
        the form $L\rightarrow{}R$, where $L$ is a single non-terminal
        symbol, and $R$ is a (possibly empty) string of terminal or
        non-terminal symbols.  Every CFG can be recognised by a
        \acronym{PDA}, and \acronyms{PDA} cannot recognise more complicated
        grammars, so \acronyms{CFG} and \acronyms{PDA} are equivalent.
    }
}

\newglossaryentry{PDA}{
    type=\acronymtype,
    name={PDA},
    description={Push-Down Automata},
    first={Push-Down Automata (PDA)}
}
\newglossaryentry{PDA glossary}{
    name={Push-Down Automata},
    description={
        A Push-Down Automation is a computational device similar to a
        \acronym{FA}, but it can additionally use a stack to store data.  A
        \acronym{PDA} can manipulate the stack during state transitions
        (adding or removing a single piece of data), and when determining
        which state transition to take it can use the piece of data on the
        top of the stack in addition to the input.  \acronyms{PDA} can
        recognise languages of the form $a^{n}b^{n}$, i.e.\ any number of
        one symbol followed by the same number of another symbol.
    }
}

>>>>

\newglossaryentry{DNSBL}{
    type=\acronymtype,
    name={DNSBL},
    description={DNS Blacklist},
    first={DNS Blacklist (DNSBL)}
}
\newglossaryentry{DNSBL glossary}{
    name={DNS Blacklist},
    description={
        A \acronym{DNSBL} is a simple collaborative anti-spam technique
        used to reject or penalise mail sent from IP addresses believed to
        be the source of large volumes of spam.  The criteria used when
        deciding if an IP address should be included vary widely between
        \acronym[s]{DNSBL}, so before using one it is very important to
        check their listing policies.  To use a \acronym{DNSBL}, Postfix
        makes a DNS request incorporating the IP address of the client; if
        the requested hostname is found the client is on the
        \acronym{DNSBL}.
    }
}

\newglossaryentry{API}{
    type=\acronymtype,
    name={API},
    description={Application Programming Interface},
    first={Application Programming Interface (API)}
}
\newglossaryentry{API glossary}{
    name={Application Programming Interface},
    description={
        One of the fundamental concepts when writing programs is the reuse
        of existing code, so that each new program does not reinvent
        existing wheels.  An \acronym{API} defines the interface provided
        to the user of the existing code, and acts as a contract between
        the user and the provider: if the user adheres to the \acronym{API}
        the provider guarantees it will work, but is free to change the
        underlying implementation if the \acronym{API} is preserved.
    }
}

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newcommand{\SMTPglossaryDescription}[0]{%
    \glsreset{SMTP}\acronym{SMTP} is the protocol used for transferring
    mail between the sending and receiving \acronym{MTA}\@.  It is a
    simple, human readable, plain text protocol, making it quite simple to
    test and debug problems with it.  A detailed description of
    \acronym{SMTP} is beyond the scope of this thesis: the original
    protocol definition is in \acronym{RFC}~821~\cite{RFC821}, which was
    superceded by \acronym{RFC}~2821~\cite{RFC2821}.%
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        \SMTPglossaryDescription{}
    }
}

\newglossaryentry{LMTP}{
    type=\acronymtype,
    name={LMTP},
    description={Local Mail Transfer Protocol},
    first={Local Mail Transfer Protocol (LMTP)}
}
\newglossaryentry{LMTP glossary}{
    name={Local Mail Transfer Protocol},
    description={
        \acronym{LMTP} is a protocol derived from \acronym{SMTP} that
        removes the need for the server to maintain a mail delivery queue,
        instead relying on the client to maintain it.  Typically the client
        is an \acronym{MTA}, and the server a delivery agent or a mail
        store.  Full details are available in~\cite{lmtp-rfc}.
    }
}

\newglossaryentry{MTA}{
    type=\acronymtype,
    name={MTA},
    description={Mail Transfer Agent},
    first={Mail Transfer Agent (MTA)}
}
\newglossaryentry{MTA glossary}{
    name={Mail Transfer Agent},
    description={
        \glsreset{MTA}A \acronym{MTA} sends and receives mail via
        \acronym{SMTP}\@.  Users submit mail to an \acronym{MTA} via their
        mail client (e.g.\ Microsoft Outlook, Thunderbird, webmail
        services); the sending \acronym{MTA} transfers the mail to the
        receiving \acronym{MTA}.
    }
}

\newglossaryentry{RFC}{
    type=\acronymtype,
    name={RFC},
    description={Request For Comments},
    first={Request For Comments (RFC)}
}
\newglossaryentry{RFC glossary}{
    name={Request For Comments},
    description={
        \glsreset{RFC}The \acronym{RFC} series is a series of proposals
        defining various protocols and file formats, e.g.\
        \acronym{SMTP}\@.  The name is somewhat misleading nowadays:
        initially the authors were asking for peer review, but these
        documents are now the de facto standards the Internet runs on.
    }
}

@@

\newglossaryentry{FQDN}{
    type=\acronymtype,
    name={FQDN},
    description={Fully Qualified Domain Name},
    first={Fully Qualified Domain Name (FQDN)}
}
\newglossaryentry{FQDN glossary}{
    name={Fully Qualified Domain Name},
    description={
        An \acronym{FQDN} is a hostname plus domain name, e.g.\ \newline{}
            example.com     is a domain name                    \newline{}
            www             is a hostname                       \newline{}
            www.example.com is a \acronym{FQDN}
    }
}

\newglossaryentry{pid}{
    type=\acronymtype,
    name={pid},
    description={Process Identifier},
    first={Process Identifier (pid)}
}
\newglossaryentry{pid glossary}{
    name={Process Identifier},
    description={
        There may be multiple copies of any program executing at any one
        time, so the program's name is not suitable as a distinguishing
        identifier; instead, each process is given a \acronym{pid} that is
        guaranteed to be unique for the lifetime of the process.  Once the
        process has completed, the pid may be reused, because they are
        drawn from a finite pool.
    }
}

\newglossaryentry{UCE}{
    type=\acronymtype,
    name={UCE},
    description={Unsolicited Commercial Email},
    first={Unsolicited Commercial Email (UCE)}
}
\newglossaryentry{UCE glossary}{
    name={Unsolicited Commercial Email},
    description={
        \acronym{UCE} is a more restrictive definition of spam than most
        people use: it only covers mail that is explicitly commercial, thus
        excluding viruses, Bayesian poisoning mails,
        backscatter\glsadd{Backscatter}, and those annoying chain letters
        you get from friends.
    }
}

\newglossaryentry{ESMTP}{
    type=\acronymtype,
    name={ESMTP},
    description={Extended SMTP},
    first={Extended SMTP (ESMTP)}
}
\newglossaryentry{ESMTP glossary}{
    name={Extended SMTP},
    description={
        \acronym{ESMTP}~\cite{RFC1869} provides a flexible mechanism for
        \acronym{SMTP} to be extended with new functionality, allowing new
        features to be tested without having to be included in the standard
        protocol.  \acronym{ESMTP} is backwards compatible with
        \acronym{SMTP}: \acronym{ESMTP} clients and servers can interact
        with \acronym{SMTP} clients and servers without difficulty.
    }
}

@@

\newglossaryentry{FA}{
    type=\acronymtype,
    name={FA},
    description={Finite Automata},
    first={Finite Automata (FA)}
}
\newglossaryentry{FA glossary}{
    name={Finite Automata},
    description={
        \acronym{FA} are computing devices that accept or recognize
        regular languages.  They lack any form of storage, so although they
        can recognise languages such as ${(ab)}^{*}$ or
        $a^{*}{(b^{*}|c)}b$, they cannot count and so cannot recognise
        languages such as $a^{n}b^{n}$.
    }
}

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        A \acronym{CFG} is a grammar in which every transition or
        production takes the form $L\rightarrow{}R$, where $L$ is a single
        non-terminal symbol, and $R$ is a (possibly empty) string of
        terminal or non-terminal symbols.  Every \acronym{CFG} can be
        recognised by a \acronym{PDA}, and \acronyms{PDA} cannot recognise
        more complicated grammars, so \acronyms{CFG} and \acronyms{PDA} are
        equivalent.
    }
}

\newglossaryentry{PDA}{
    type=\acronymtype,
    name={PDA},
    description={Push-Down Automata},
    first={Push-Down Automata (PDA)}
}
\newglossaryentry{PDA glossary}{
    name={Push-Down Automata},
    description={
        \glsreset{PDA}A \acronym{PDA} is a computational device similar to
        a \acronym{FA}, but it can additionally use a stack to store data.
        A \acronym{PDA} can manipulate the stack during state transitions
        by adding or removing a single piece of data, and when determining
        which state transition to take, the piece of data on the top of the
        stack can be used in addition to the input.  In addition to the
        languages recognised by \acronyms{FA}, \acronyms{PDA} can recognise
        languages of the form $a^{n}b^{n}$, i.e.\ any number of one symbol
        followed by the same number of another symbol.
    }
}

### 2009-04-22 doc/logparser-background.tex
Corrections after proofreading chapters 6 & 7 and the appendices.

    \item \acronym{SPF}~\cite{openspf}.  \acronym{SPF} records specify
        which mail servers are allowed to send mail using sender addresses
        from a particular domain.  The intention is to reduce spam from
        faked sender addresses, backscatter~\cite{postfix-backscatter}, and
        Joe~jobs~\glsadd{Joe-job}.  There has been considerable resistance
        to \acronym{SPF} because it breaks or vastly complicates some
        commonly-used features of \acronym{SMTP}, e.g.\ forwarding mail
        from one domain to another when a user moves.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote IP
        address)}\newline{} has not been seen before; on second and
        subsequent delivery attempts from that triple the mail will be
        accepted.  This blocks spam from some senders because maintaining a
        list of failed addresses and retrying after a temporary failure is
        uneconomical for a spam sender, but a legitimate mail server must
        retry deliveries that fail temporarily.  Sadly spam senders are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided \acronym{SMTP} server from a
        compromised machine on the ISP's network.  Greylisting will slowly
        become less effective as spam senders adapt, but it does block a
        large percentage of spam mail at the moment; the most effective
        restrictions from the \numberOFlogFILES{} log files used when
        generating the results in \sectionref{Evaluation} are shown in
        \tableref{Most effective Postfix restrictions}.  The table shows
        that greylisting is worth using at the moment, particularly when
        you take into account its position as the final restriction that a
        mail must overcome in the configuration used on the mail server
        that generated the log files: on that server greylisting only takes
        effect for mails that have passed all other restrictions.  Some
        problems may be encountered when using greylisting: some servers
        fail to retry after a temporary failure, or legitimate mail may be
        delayed, particularly when coming from a pool of servers.

@@

    \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
        perform tests on features of the delivery attempt (e.g.\ IP
        address, sender address), incrementing or decrementing a score
        based on the results; if the final score is higher than a threshold
        the delivery attempt is rejected.  The administrator must manually
        whitelist clients if they are to bypass a Postfix restriction,
        whereas using a threshold that requires a delivery attempt to hit
        several scored restrictions will allow mail from clients that would
        be rejected by a binary restriction.

>>>>

    \item \acronym{SPF} checks:
        \urlLastChecked{http://www.openspf.org/}{2009/04/21}.
        \acronym{SPF} records specify which mail servers are allowed to
        send mail using sender addresses from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter}, and
        Joe~jobs~\glsadd{Joe-job}.  There has been considerable resistance
        to \acronym{SPF} because it breaks or vastly complicates some
        commonly-used features of \acronym{SMTP}, e.g.\ forwarding mail
        from one domain to another when a user moves.

    \item Greylisting,
        \urlLastChecked{http://www.greylisting.org/}{2009/04/21}, is a
        technique that temporarily rejects a delivery attempt when the
        triple of \newline{} \tab{}\texttt{(sender address, recipient
        address, remote IP address)}\newline{} has not been seen before; on
        second and subsequent delivery attempts from that triple the mail
        will be accepted.  This blocks spam from some senders because
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spam sender, but a
        legitimate mail server must retry deliveries that fail temporarily.
        Sadly spam senders are using increasingly complex and well written
        programs to distribute spam, frequently using an ISP provided
        \acronym{SMTP} server from a compromised machine on the ISP's
        network.  Greylisting will slowly become less effective as spam
        senders adapt, but it does block a large percentage of spam mail at
        the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used when generating the results in
        \sectionref{Evaluation} are shown in \tableref{Most effective
        Postfix restrictions}.  The table shows that greylisting is worth
        using at the moment, particularly when you take into account its
        position as the final restriction that a mail must overcome in the
        configuration used on the mail server that generated the log files:
        on that server greylisting only takes effect for mails that have
        passed all other restrictions.  Some problems may be encountered
        when using greylisting: some servers fail to retry after a
        temporary failure, or legitimate mail may be delayed, particularly
        when coming from a pool of servers.

@@

    \item Scoring systems such as postfwd,
        \urlLastChecked{http://postfwd.org/}{2009/04/21}, perform tests on
        features of the delivery attempt (e.g.\ IP address, sender
        address), incrementing or decrementing a score based on the
        results; if the final score is higher than a threshold the delivery
        attempt is rejected.  The administrator must manually whitelist
        clients if they are to bypass a Postfix restriction, whereas using
        a threshold that requires a delivery attempt to hit several scored
        restrictions will allow mail from clients that would be rejected by
        a binary restriction.

### 2009-04-22 doc/logparser-conclusion.tex
Corrections after proofreading chapters 6 & 7 and the appendices.

Parsing Postfix log files appears at first sight to be an almost trivial
task, especially if one has previous experience in parsing log files, but
it turns out to be a much more taxing project than initially expected.  The
variety and breadth of log lines produced by Postfix is quite surprising,
because a quick survey of sample log files gives the impression that the
number of distinct log line variants is quite small; this mistaken
impression is because of the uneven distribution exhibited by log lines
produced in normal operation, vividly illustrated by \graphref{rule hits
graph}.

Given the diverse nature of Postfix log lines, and the ease with which
administrators can cause new log lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Providing a
tool to ease the generation of regexes from unrecognised log lines should
greatly help users to extend their ruleset to recognise previously
unrecognised log lines (\sectionref{creating new rules in implementation}).
The architecture developed for this project not only makes it trivial to
recognise new inputs, adding new actions requires the minimum effort
possible, allowing the implementer to focus on correctly implementing their
new action.  The separation of the architecture into rules, actions, and
framework (\sectionref{parser design}) is unusual, partly because the three
are separated so completely.  Although parsers are often divided into
separate source code files (the combination of lex \&
yacc~\cite{lex-and-yacc-book} being a common example), the parts are
usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.

The flow of control in this architecture is quite different from most other
architectures.  Most traditional parsers have a current state: each state
has a fixed set of acceptable next states, and processing is determined by
the state transition that takes place; unacceptable state transitions cause
parsing to fail.  For example, when a \textbf{C} parser sees the keyword
\textbf{for} it expects to immediately see a left parenthesis, with any
other input causing parsing to fail (comments will already have been
removed by the preprocessor).  This architecture is different: the rule
that recognises the input dictates the action that will be invoked.  Rule
conditions (\sectionref{rule conditions in architecture}) allow a stateful
implementation of this architecture, where the list of rules used to
recognise an input is constrained by the parser's current state, but the
recognising rule still dictates the action that is invoked, and whether
directly or indirectly, the next state.

XXX SHOULD THIS NEXT PARAGRAPH BE IN THE PAST TENSE\@?

The real difficulties arose once the parser was successfully dealing with
90\% of the log lines, because most of the irregularities and complications
explained in \sectionref{complications} only became apparent when the vast
majority of log lines were being parsed successfully.  Adding new rules to
deal with the numerous infrequently occurring log line variants is a simple
if tiresome task, whereas dealing with mails that are missing information
or where Postfix's actions are not being correctly reconstructed is much
more grueling.  Trawling through log files, searching for something out of
the ordinary, possibly hundreds or even thousands of log lines away from
the last occurrence of the queueid for the mail in question, is extremely
time consuming and quite error prone.  Sometimes the task is not to spot
the unusual log line, but to spot that a log line that is normally present
is missing, i.e.\ to realise that one log line amongst thousands is absent.
In all cases the evidence must be used to construct a hypothesis to explain
the irregularities, and that hypothesis has to be tested in \parsername{};
if successful, the parser will be modified to deal with the irregularities,
without adversely affecting the existing parsing.  The complications
described in \sectionref{complications} are described in the order they
were solved in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they were the first
complications to be dealt with.

XXX WRITE ABOUT EFFICIENCY\@.

A parser's ability to correctly parse its input is extremely important; the
parser's coverage of \numberOFlogFILES{} log files is discussed in
\sectionref{parsing coverage}.  Both its success at recognising individual
log lines (\sectionref{log-lines-covered}) and its correctness in
reconstructing each mail's journey through Postfix
(\sectionref{mails-covered}) are described in detail, including the results
of manually verifying the correct parsing of a subset of a log file.
Experience implementing \parsername{} shows that full input coverage is
relatively easy to achieve with this architecture, and that with enough
time and effort a full understanding of the input is possible.  Postfix log
files would require substantial time and effort to correctly parse
regardless of the architecture used; this architecture enables an iterative
approach to be used (similar to Stepwise
Refinement~\cite{stepwise-refinement}), as is practiced in many other
software engineering disciplines.

This architecture's greatest strength is the ease with which
implementations of it can be adapted to deal with new requirements and
inputs.  Parsing a variation of an existing input is a trivial task: simply
modify an existing rule or add a new rule and the task is complete.
Parsing a new category of input is achieved by writing a new action and
appropriate rules; quite often the new action will not need to interact
with existing actions, but when interaction is required the framework
provides the necessary facilities.  The decoupling of rules from actions
allows different sets of rules to be used with one set of actions, e.g.\ a
parser might have actions to process versions one and two of a file format;
by choosing the appropriate ruleset the parser will parse version one, or
version two, or both versions.  Decoupling also allows other approaches to
rule management, as discussed in \sectionref{adding new rules in
architecture}.  The architecture makes it possible to apply commonly used
programming techniques (such as object orientation, inheritance,
composition, delegation, roles, modularisation, or closures) when designing
and implementing a parser, simplifying the process of working within a team
or when developing and testing additional functionality.  This architecture
is ideally suited to parsing inputs that are not fully understood or do not
follow a fixed grammar: the architecture warns about unparsed inputs and
other errors, but continues parsing as best it can, allowing the developer
of a new parser to decide which deficiencies are most important and require
attention first, rather than being forced to fix the first error that
arises.

The data gathered by \parsername{} provides the foundation for the future
of this project: applying machine-learning algorithms to the data to
analyse and optimise the set of anti-spam defences in use, followed by
identifying patterns in the data that could be used to write new anti-spam
techniques to recognise and reject spam rather than accepting it.  The
parser provides the data in a normalised form that is far easier to use as
input to new or existing implementations of machine-learning algorithms
than trying to adapt each algorithm to extract data directly from the log
files.  New policy servers (\sectionref{policy servers}), written to
implement new anti-spam measures, could be tested or trained by using the
data to simulate mail delivery attempts; this would allow simple, fast,
reproducible testing, without the risk of adversely affecting a production
mail server.  Development of \parsername{} is complete, i.e.\ it correctly
parses Postfix log files, and in future it will require only maintenance;
however, one avenue of future development under consideration is to extend
it to parse non-Postfix log lines, e.g.\ SpamAssassin or Amavisd-new log
lines.  \parsername{} can easily be extended to do this, but it lacks a
method of associating the non-Postfix log lines with the existing data
structures and state tables, so that all of the data for a mail delivery
attempt can be stored together.

The Postfix log file parser based on this architecture provides a basis for
systems administrators to monitor the effectiveness of their anti-spam
measures and adapt their defences to combat new techniques used by those
sending spam.  \parsername{} is a fully usable application, built to
address a genuine need, rather than a proof of concept whose sole purpose
is to illustrate a new idea; it deals with the oddities and difficulties
that occur in the real world, rather than a clean, idealised scenario
developed to showcase the best features of a new approach.

>>>>

Parsing Postfix log files appears at first sight to be an uncomplicated
task, especially if one has previous experience in parsing log files, but
it turns out to be a much more taxing project than initially expected.  The
variety and breadth of log lines produced by Postfix is quite surprising,
because a quick survey of sample log files gives the impression that the
number of distinct log line variants is quite small; this mistaken
impression comes from the uneven distribution exhibited by log lines
produced in normal operation, vividly illustrated in \graphref{rule hits
graph}.  Given the diverse nature of Postfix log lines, and the ease with
which administrators can cause new log lines to be logged
(\sectionref{postfix background}), enabling users to easily extend the
parser to deal with new log lines is a design imperative
(\sectionref{parser design}).  Providing a tool to ease the generation of
regexes from unrecognised log lines (\sectionref{creating new rules in
implementation}) should greatly help users who need to extend their ruleset
to recognise previously unrecognised log lines.

This architecture's greatest strength is the ease with which parsers based
on it can be adapted to deal with new requirements and inputs.  Parsing a
variation of an existing input is a trivial task: simply modify an existing
rule or add a new rule and the task is complete.  Parsing a new category of
input is achieved by writing a new action and a rule for each input
variant; quite often the new action will not need to interact with existing
actions, but when interaction is required the framework provides the
necessary facilities.  The architecture imposes very little red tape when
writing new actions, allowing the implementer to focus their time and
energy on correctly implementing their new action (\sectionref{actions in
architecture}).  The separation of the architecture into rules, actions,
and framework (\sectionref{parser design}) is unusual, partly because the
three are separated so completely.  Although parsers are often divided into
separate source code files (the combination of lex \&
yacc~\cite{lex-and-yacc-book} being a common example), the parts are
usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  This separation
enables the optimisations discussed in \sectionref{parser efficiency}, and
it also allows different approaches to ruleset management, e.g.\ using
machine learning techniques to seamlessly create or alter rules to
recognise new inputs (\sectionref{adding new rules in architecture}).  The
decoupling of rules from actions allows different sets of rules to be used
with one set of actions, e.g.\ a parser might have actions to process
versions one and two of a file format; by choosing the appropriate ruleset
the parser will parse version one, or version two, or both versions.  A
general purpose framework can be written, so that writing a parser just
requires writing actions and rules.  The architecture makes it possible to
apply commonly used programming techniques (such as object orientation,
inheritance, composition, delegation, roles, modularisation, or closures)
when designing and implementing a parser, simplifying the process of
working within a team or developing and testing additional functionality.
This architecture is ideally suited to parsing inputs that are not fully
understood or do not follow a fixed grammar: the architecture warns about
unrecognised inputs and errors encountered by actions, but continues
parsing as best it can, allowing the developer of a new parser to decide
which deficiencies are most important and require attention first, rather
than being forced to fix the first error that arises.

The flow of control in this architecture is quite different from most other
architectures.  Most traditional parsers have a current state: each state
has a fixed set of acceptable next states, processing is determined by the
state transition that takes place, and unacceptable state transitions cause
parsing to fail.  This architecture is different: the rule that recognises
the input dictates the action that will be invoked.  Rule conditions
(\sectionref{rule conditions in architecture}) enable stateful parsing,
where the list of rules used to recognise an input is constrained by the
parser's current state, but the recognising rule still dictates the action
that is invoked and, whether directly or indirectly, the next state.

When writing \parsername{}, the real difficulties arose once the parser was
successfully recognising almost all of the log lines, because most of the
irregularities and complications explained in \sectionref{complications}
started to become apparent then.  Adding new rules to deal with numerous
infrequently occurring log line variants was a simple if tiresome task,
whereas dealing with mails that were missing information or where Postfix's
actions were not being correctly reconstructed was much more grueling.
Trawling through log files was extremely time consuming and quite error
prone, searching for something out of the ordinary that might help diagnose
the problem, and eventually finding it --- sometimes hundreds or even
thousands of log lines away from the last occurrence of the queueid for the
mail in question.  Sometimes the task was not to identify the unusual log
line, but to spot that a log line normally present was missing, i.e.\ to
realise that one log line amongst thousands was absent.  In all cases the
evidence was used to construct a hypothesis to explain the irregularities,
and that hypothesis was tested in \parsername{}; if successful, the parser
was modified to deal with the irregularities, without adversely affecting
existing parsing.  The complications documented in
\sectionref{complications} are presented in the order they were solved in,
and that order closely resembles the frequency in which they occur; the
most frequently occurring complications dominate the warning messages
produced, and so naturally they were the first complications to be dealt
with.

\parsername{} is not merely a proof of concept: it is intended to be used
for parsing real-world log files from production mail servers, and the
resulting data used to improve anti-spam defences.  This means that
efficiency is important: parsing must complete in an reasonable period of
time, so that the results can be used in a timely manner.  \parsernames{}
efficiency is evaluated in \sectionref{parser efficiency}, where
optimisations and the effect they have are explored.

A parser's ability to correctly parse its inputs is extremely important;
\parsernames{} coverage of \numberOFlogFILES{} log files is discussed in
\sectionref{parsing coverage}.  Both its success at recognising individual
log lines  and its correctness in reconstructing each mail's journey
through Postfix are described in detail, including the results of manually
verifying that a randomly selected portion of a log file was correctly
parsed.  Experience implementing \parsername{} shows that full input
coverage is relatively easy to achieve with this architecture, and that
with enough time and effort a full understanding of the input is possible.
Postfix log files would require substantial time and effort to correctly
parse regardless of the architecture used; this architecture enables an
iterative approach to be used~\cite{stepwise-refinement}, as is practiced
in many other software engineering disciplines.

The data gathered by \parsername{} provides the foundation for the future
of this project: using machine-learning algorithms to analyse the data and
optimise the set of anti-spam defences in use, followed by identifying
patterns in the data that could be used to write new anti-spam techniques
to recognise and reject spam rather than accepting it.  The database
provides the data in a normalised form that is far easier to use as input
to new or existing implementations of machine-learning algorithms than
trying to adapt each algorithm to extract data directly from log files.
New policy servers, written to implement new anti-spam measures, can be
tested or trained by using the collected data to simulate mail delivery
attempts; this would allow simple, fast, reproducible testing, without the
risk of adversely affecting a production mail server.  Development of
\parsername{} is finished, i.e.\ it correctly parses Postfix log files, and
in future it will only require maintenance; however, one avenue of future
development under consideration is to extend it to parse non-Postfix log
lines, e.g.\ SpamAssassin or Amavisd-new log lines.  \parsername{} can
easily be extended to do this, but it requires a method of associating the
non-Postfix log lines with the existing data structures and state tables,
so that all of the data for a mail delivery attempt can be stored together.

\parsername{} provides a basis for systems administrators to monitor the
effectiveness of their anti-spam measures and adapt their defences to
combat new techniques used by those sending spam.  \parsername{} is a fully
usable application, built to address a genuine need, rather than a proof of
concept whose sole purpose is to illustrate a new idea; it deals with the
oddities and difficulties that occur in the real world, rather than a
clean, idealised scenario developed to showcase the best features of a new
approach.
